---
title: "Sentiment Analysis of COVID-19 Tweets"
output:
  pdf_document:
    latex_engine: xelatex
date: "2024-11-11"
---
Introduction: 
The purpose of this assignment is to conduct sentiment analysis. I decided to analyze a collection of tweets to explore and quantify the emotional tone of social media interactions during COVID-19. Sentiment analysis is a natural language processing technique used to classify text by identifying and categorizing emotions or attitudes conveyed within the language. I will beapplying different sentiment lexicons — namely, AFINN, Bing, and NRC. 

The AFINN lexicon assigns sentiment scores, allowing a calculation of cumulative sentiment per tweet, while the Bing lexicon categorizes words simply as "positive" or "negative." The NRC lexicon provides a broader emotional classification, encompassing various sentiment categories.

```{r}
library(tidytext)

get_sentiments("afinn")
get_sentiments("bing")
get_sentiments("nrc")
```


```{r}
# Load Libraries 
library(rtweet)
library(tidytext)
library(dplyr)
library(ggplot2)

```

```{r}
# Load the dataset
tweets <- read.csv("/Users/leslietavarez/Downloads/covid19_tweets.csv")

# Preview the data
head(tweets)
```
```{r}
# Check structure and add unique IDs to each tweet row
tweets_clean <- tweets %>%
  mutate(tweet_id = row_number()) %>% # Creates a unique ID for each tweet row
 select(tweet_id, text)

# Tokenize text by individual tweet and remove stop words
tweets_words <- tweets_clean %>%
  unnest_tokens(word, text) %>%
  anti_join(stop_words, by = "word")

head(tweets_clean)
```
The histogram from the AFINN lexicon reveals a nearly symmetric distribution around zero, indicating a balanced mix of positive and negative sentiment in the COVID-19 tweets.

```{r}
# Perform sentiment analysis using AFINN lexicon
tweets_sentiment <- tweets_words %>%
  inner_join(get_sentiments("afinn"), by = "word") %>%
  group_by(tweet_id) %>%  # assuming 'tweet_id' is a unique identifier for each tweet
  summarize(sentiment_score = sum(value, na.rm = TRUE))

#View results
tweets_sentiment

# Histogram of sentiment scores
ggplot(tweets_sentiment, aes(x = sentiment_score)) +
  geom_histogram(binwidth = 1, fill = "skyblue", color = "black") +
  labs(title = "Distribution of Sentiment Scores",
       x = "Sentiment Score",
       y = "Count of Tweets") +
  theme_minimal()
```
The positive sentiment is surprisingly strong, with more positive than negative tweets overall. However, there’s a noticeable mix of emotions—anticipation, fear, sadness, and trust all show up frequently. It’s interesting to see so many positive tweets, as I expected more negativity considering how hard the pandemic was for many, with people losing loved ones and facing uncertainty. This mix of sentiments really captures the complex emotions people felt during that time.

```{r}
# Perform sentiment analysis using NRC lexicon
tweets_sentiment_nrc <- tweets_words %>%
  inner_join(get_sentiments("nrc"), by = "word") %>%
  group_by(tweet_id, sentiment) %>%  # Group by tweet and sentiment type
  summarize(sentiment_count = n(), .groups = 'drop')  # Count occurrences of each sentiment per tweet

# View the results
head(tweets_sentiment_nrc)
tweets_sentiment_nrc

# Heatmap of sentiment counts
ggplot(tweets_sentiment_nrc, aes(x = factor(tweet_id), y = sentiment, fill = sentiment_count)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "blue") +
  labs(title = "Heatmap of Sentiments per Tweet",
       x = "Tweet ID",
       y = "Sentiment") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


# Summarize total counts per sentiment
sentiment_summary <- tweets_sentiment_nrc %>%
  group_by(sentiment) %>%
  summarize(total_count = sum(sentiment_count))

# Bar plot
ggplot(sentiment_summary, aes(x = sentiment, y = total_count, fill = sentiment)) +
  geom_bar(stat = "identity") +
  labs(title = "Total Sentiment Counts Across All Tweets",
       x = "Sentiment",
       y = "Total Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The Bing lexicon analysis reveals a predominance of negative sentiment over positive sentiment in the COVID-19 tweets.This reflects the challenging aspects of the pandemic.However, the presence of positive sentiment shows that some tweets conveyed hope or moments of appreciation.


```{r}
# Perform sentiment analysis using the Bing lexicon
tweets_sentiment_bing <- tweets_words %>%
  inner_join(get_sentiments("bing"), by = "word") %>%  # Join with Bing lexicon
  group_by(tweet_id, sentiment) %>%                    # Group by tweet and sentiment type
  summarize(sentiment_count = n(), .groups = 'drop')   # Count positive/negative words for each tweet

# View the results
head(tweets_sentiment_bing)

# Summarize total positive and negative counts
sentiment_summary <- tweets_sentiment_bing %>%
  group_by(sentiment) %>%
  summarize(total_count = sum(sentiment_count))

# Bar plot for total positive vs. negative sentiment
ggplot(sentiment_summary, aes(x = sentiment, y = total_count, fill = sentiment)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("positive" = "skyblue", "negative" = "salmon")) +
  labs(title = "Total Positive vs. Negative Sentiments",
       x = "Sentiment",
       y = "Total Count of Words") +
  theme_minimal()

```
Conclusion: 

The Bing lexicon appears more negative than others because it classifies words like "quarantine" or "isolation" as inherently negative, even though they might be neutral in context. Bing’s binary approach exaggerates negativity, especially in difficult topics like COVID-19. In contrast, NRC captures a broader range of emotions, leading to a more balanced distribution, while AFINN’s numerical scoring system allows for more nuance, offsetting mild negativity with extreme positivity for a more balanced sentiment analysis.